\propertitle

### 2019-12-02 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-02T17:12:16+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The first and most important assumption of this kind of maximum-entropy
application in neuroscience can be expressed in three equivalent ways:
\begin{enumerate*}[(\roman*)]
\item the order of the time bins has no relevance; \item the frequencies of
  the activity during a recording are informationally sufficient for
  inferences about unrecorded times; \item our degree of belief about the
  sequence of activities is invariant under permutations of the time bins,
  that is, exchangeable \citep[\chap~4]{bernardoetal1994_r2000}{dawid2013}.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
\citep{definetti1930,hewittetal1955} is that the probability of observing a
sequence of activities $(\yA_{1},\dotsc,\yA_{T})$ with frequencies
$(\yFF_{\yA})$ is a mixture of terms
\begin{equation}
  \label{eq:mixture_from_definetti}
%  \exp[-T\;\sh(\yF;\ynu)] \quad
%  \pxxx[(\yA_{1},\dotsc,\yA_{T})] \propto
  \prod_{\yA=1}^{N}{\nu_{\yA}}^{T\,\yFF_{\yA}} \quad
  \text{with weights }q(\ynu),
\end{equation}
where $\ynu\defd (\nu_{\yA})$ represents the long-run frequency
distribution for the activity. % Other maximum-entropy models are also used
% in neuroscience, such as kinetic Ising models
% \cite{kawasaki1972,mezardetal2011,battistinetal2015}, based on the slightly
% less drastic assumption of invariance under permutations of sequences
% having the same starting state and the same transition counts
% \citep{diaconisetal1980c,geetal2012}.
Note that the maximum-entropy method in statistical physics is meant to be
used not with this assumption, but with averages across \emph{repetitions}
of the same experiment: its constraints represent macroscopically
\emph{reproducible} quantities and kinematics, even in non-equilibrium
\citep{jaynes1957,guntonetal1983,tikochinskyetal1984,grandy1988,berryetal1988,deroecketal2005_r2006,touchette2009}.

The second assumption is specific to our application: all subsets of $\yA$
neurons in the larger population are equally likely to give rise to total
activity $\yA$, at every time bin. A mathematically equivalent assumption
is that $n$ neurons are sampled anew at each time bin. This assumption
allows us to factorize the joint probability of the two sequences of
activities and apply the sampling
formula~\eqref{eq:conditional_hypergeometric} at each bin:
\begin{equation}
  \label{eq:assume_sampled_anew}
  \pxx[(\ya_{1},\dotsc,\ya_{T}) \| (\yA_{1}, \dotsc, \yA_{T}) ]
  = \prod_{t=1}^{T} \pxx(\ya_{t} \| \yA_{t})
  = \prod_{t=1}^{T} \yG_{\ya_{t}\yA_{t}}.
\end{equation}

@@

From these two assumptions the probability calculus allows us to calculate
the probability $\pxx(\yF \| \yf )$ for the frequency distribution $\yF$
given the observed frequency distribution for the sample $\yf$. The result
is a mixture of terms involving the joint frequencies of total activities
of sample and larger population $J_{\ya,\yA}$, multiplied by the number of
time bins $T$:
\begin{equation}
  \label{eq:prob_joint_frequencies}
%\pxx[\yF \| \yf ] \propto
 \sum_{\bm{J}} \prod_{\ya,\yA}\bigl(\yG_{\ya\yA}\;\nu_{\yA}\bigr)^{T\,J_{\ya,\yA}}
  \qquad
  \text{with}\quad \tsum_{\yA}J_{\ya,\yA}=\yff_{\ya},
  \quad \tsum_{\ya}J_{\ya,\yA}=\yFF_{\yA},
\end{equation}
with weights $q(\ynu)$. The maximum-entropy method approximates $T$ with
infinity, thus treating the probability above as a delta. The result is
that
\begin{enumerate*}[label=(\roman*)]
\item the frequency distribution $\yF$ for the experiment and the long-run
  one $\ynu$ are constrained to be equal,
\item the frequency of the sample activity $\ya$ conditional on the
  larger-population activity $\yA$ is constrained to equal the
  hypergeometric distribution $\yG_{\ya\yA}$.
\end{enumerate*}
Without this approximation the latter conditional frequency can differ from
$\yG_{\ya\yA}$, and this is expected if the number of bins $T$ is not much
larger than the number of possible joint possibilities, roughly
$n\times N$.

The final assumption in the maximum-entropy method regards the form of the
probability $\pxx(\ynu)$ in the expressions above.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-02T22:37:28+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The first and most important assumption of this kind of maximum-entropy
applications in neuroscience can be expressed in two equivalent ways:
\begin{enumerate*}[(\roman*)]
  % \item the order of the time bins has no relevance;
\item the frequencies of the activity during a recording are
  informationally sufficient for inferences about unrecorded times; \item
  our degree of belief about the sequence of activities is exchangeable
  \citep[\chap~4]{bernardoetal1994_r2000}{dawid2013}, that is, invariant
  under permutations of the time bins.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
\citep{definetti1930,hewittetal1955} is that the probability of observing a
sequence of activities $(\yA_{1},\dotsc,\yA_{T})$ with frequencies
$(\yFF_{\yA})$ is a mixture of terms
\begin{equation}
  \label{eq:mixture_from_definetti}
%  \exp[-T\;\sh(\yF;\ynu)] \quad
%  \pxxx[(\yA_{1},\dotsc,\yA_{T})] \propto
  \prod_{\yA=1}^{N}{\nu_{\yA}}^{T\,\yFF_{\yA}} \quad
  \text{with weights }q(\ynu),
\end{equation}
where $\ynu\defd (\nu_{\yA})$ represents the long-run frequency
distribution for the activity. % Other maximum-entropy models are also used
% in neuroscience, such as kinetic Ising models
% \cite{kawasaki1972,mezardetal2011,battistinetal2015}, based on the slightly
% less drastic assumption of invariance under permutations of sequences
% having the same starting state and the same transition counts
% \citep{diaconisetal1980c,geetal2012}.
Note that the maximum-entropy method in statistical physics is meant to be
used not with this assumption, but with averages across \emph{repetitions}
of the same experiment, because its constraints represent macroscopically
\emph{reproducible} quantities and kinematics, even in non-equilibrium
\citep{jaynes1957,guntonetal1983,tikochinskyetal1984,grandy1988,berryetal1988,deroecketal2005_r2006,touchette2009}.

The second assumption is specific to our application: all subsets of $\yA$
neurons in the larger population are equally likely to give rise to total
activity $\yA$, at every time bin. A mathematically equivalent assumption
is that $n$ neurons are sampled anew at each time bin. This assumption
allows us to factorize the joint probability of the two sequences of
activities and apply the sampling
relation~\eqref{eq:conditional_hypergeometric} of
appendix~\ref{sec:derivation_maxent_fullnet} at each bin:
\begin{equation}
  \label{eq:assume_sampled_anew}
  \pxx[(\ya_{1},\dotsc,\ya_{T}) \| (\yA_{1}, \dotsc, \yA_{T}) ]
  = \prod_{t=1}^{T} \pxx(\ya_{t} \| \yA_{t})
  = \prod_{t=1}^{T} \yG_{\ya_{t}\yA_{t}}.
\end{equation}

@@

From these two assumptions the probability calculus allows us to derive not
only the probability for the frequency distribution $\yF$, but also that
for the joint frequencies of total activities of sample and larger
population, $J_{\ya,\yA}$, conditional on the observed frequency
distribution for the sample $\yf$. The result is again a mixture of terms
\begin{equation}
  \label{eq:prob_joint_frequencies}
%\pxx[\yF \| \yf ] \propto
  \prod_{\ya,\yA}\bigl(\yG_{\ya\yA}\;\nu_{\yA}\bigr)^{T\,J_{\ya,\yA}}
  \qquad
  \text{with } \tsum_{\yA}J_{\ya,\yA}=\yff_{\ya}
  %\quad \tsum_{\ya}J_{\ya,\yA}=\yFF_{\yA},
\end{equation}
with weights $q(\ynu)$.

The maximum-entropy method approximates $T$ in the probability above with
infinity, thus treating it as a delta. This leads to two approximate equalities:
\begin{enumerate*}[label=(\roman*)]
\item of the recorded and long-run frequency distributions $\yF$ and
  $\ynu$, and therefore also of their probability distributions,
\item of the hypergeometric distribution $\yG_{\ya\yA}$ and the frequency
  of the sample activity $\ya$ conditional on the larger-population
  activity $\yA$.
\end{enumerate*}
Without this approximation the latter conditional frequency can differ from
$\yG_{\ya\yA}$ -- and this is expected if the number of bins $T$ is not much
larger than the number of possible joint possibilities, roughly
$n\times N$.

The final assumption of the maximum-entropy method concerns the form of the
probability density $\pxx(\ynu)$ in the expressions above. It is taken to
be either an \enquote{entropic prior}
\citep{skillingetal1984,rodriguez1991,neumann2007}, which partially takes
into account the multiplicity of long-run frequencies, or a prior for a
model with sufficient statistics, which expresses the sufficiency of the
model under study \citep{portamana2017}. This assumption is crucial for
obtaining the maximum-entropy method with the Shannon entropy. A different
density would lead to alternative maximum-entropy methods, for example
based on the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}.
